---
title: "MIACCrossCountry"
output: html_notebook
---

```{r}
library(rvest)
library(dplyr)
library(magrittr)
library(tidyverse)
library(jsonlite)
library(stringr)
library(xml2)
library(XML)
library(rvest)
library(tidyverse)
library(httr)
library(httpuv)
library(RCurl)
library(stringr)
library(lubridate)
```

Obtain list of urls for Minnesota college cross country meets from tfrrs (2012 - 2017).
```{r}
#Scrape the data from this page: https://www.tfrrs.org/results_search.html?page=0&title=1&go=1&state=mn&sport=xc to get links to the individual result links then scrap those

resultLinks <- data.frame()
resultList <- read_html("https://www.tfrrs.org/results_search.html?page=0&title=1&go=1&state=mn&sport=xc")

# list of meet links
links <- resultList %>% 
  html_nodes("#results_search a")

# convert each link to string and add to resultLinks dataframe
for(i in 1:length(links)) {
  
  link <- links[i] %>% 
    html_attrs() %>% 
    toString()
 
  df <- tibble(link)
  resultLinks <- rbind(resultLinks,df)
}

# add appropriate prefix for links, convert to character
resultLinksClean <- resultLinks %>% 
  mutate(link = gsub("(//www)","https://www",link),
         link = as.character(link))

# write result links list to csv
write.csv(resultLinksClean,"/Users/hannahsonsalla/Desktop/Math/DataScience/Project/resultLinks.csv")

```

Data scraping: Scrape athlete data from each race using url list.
```{r}
resultLinks <- read.csv("/Users/hannahsonsalla/Desktop/Math/DataScience/Project/resultLinks.csv")

# MIAC Cross Country Championships
#resultLinks <- read_html("https://www.tfrrs.org/results/xc/6155.html")
#resultLinks <- data.frame()

resultLinks <- resultLinks %>%
  mutate(link = as.character(link))

minnesota_xc<-data.frame()

# Loop through every url in result links url list
for (url in resultLinks$link) {
    
    # Read the url page
    page<- read_html(url)
    
    #Retrieve meet name
    meet_name <-
      page %>%
      html_nodes(".title a") %>%
      html_text()
    
    #Retrieve meet date
    meet_date <-
      page %>%
      html_nodes(".datelocation li:nth-child(1)") %>%
      html_text()
    meet_date_short<-substr(meet_date,start=6,stop=14)
    
    #Retrieve meet location
    meet_location <-
      page %>%
      html_nodes(".datelocation li:nth-child(2)") %>%
      html_text()
    meet_location_short<-substr(meet_location,start=12,stop=(nchar(meet_location) - 4))
    meet_location_short = gsub("(\n)|(\nCity Not Supplied,)",", ",meet_location_short)
    
    #Retrieve start time
    start_time <-
      page %>%
      html_node("div:nth-child(12) span") %>%
      html_text()
    start_time = gsub("(\n)*(\t)*","",start_time)
    
    race <- page %>% 
      html_nodes("a+ div div")
    
    # Get list of tables on page
    tables <- page %>% 
      html_nodes("table")
    
    #Last table on page contains womens results
    lastTableLocation <- length(tables)
    
    # Every other table contains individual results
    tableList = c(lastTableLocation)
    tableRef = lastTableLocation - 2
    while(tableRef >= 4) {
      tableList=c(tableRef,tableList)
      tableRef = tableRef - 2
    }
    
    # Loop through individual result tables 
    for(t in (1:length(tableList))) { 
      
      # Retrieve race group (Women's 6k, Men's 8k, etc.)
      race_group=""
      if (length(race[t] %>% html_text()>0)) {
        race_group = race[t] %>% html_text()
        race_group = gsub("(\n)*(\t)*","",race_group)
      } 
    
      lastTableDivs <- tables[tableList[t]] %>% 
        html_nodes("tr") %>% 
        html_nodes("td") %>% 
        html_nodes("div")
      
      #Need to change this to only loop through first 7 td divs in the table
      tableRows <- tables[tableList[t]] %>% 
        html_nodes("tr")
      
      for(row in (1:length(tableRows))) {
        
        tableDivs <- tableRows[row] %>% 
          html_nodes("td")  %>% 
          html_nodes("div")
        
        # As long as table not empty
        if(length(tableDivs)>0) {
        
         # Retrieve place
          place <- tableDivs[1] %>%
            html_text()
          
          # Retrieve and inspect full names
          name <- tableDivs[2] %>%
            html_text()
      
          # Retrieve and inspect class year
          class_year <- tableDivs[3] %>%
            html_text()
      
          # Retrieve and inspect team
          team <- tableDivs[4] %>%
            html_text()
      
          # Retrieve and inspect avg. mile
          avg_mile <- tableDivs[5] %>%
            html_text()
          
          # Retrieve time
          time <- tableDivs[6] %>%
            html_text()
          
          # Retrieve score
          score <- tableDivs[7] %>%
            html_text()
          
          
          meet_info <- 
            tibble(meet_name = meet_name, meet_date = meet_date_short, meet_location = meet_location_short, start_time=start_time, race_group=race_group, place = place, name = name, class_year = class_year, team = team, avg_mile = avg_mile, time = time, score = score, url=url)
        
          #store the temp values into the master dataframe
          minnesota_xc<-rbind(minnesota_xc, meet_info)
        }
      }
    }
    print(url)
} 

head(minnesota_xc)

#write.csv(minnesota_xc,"/Users/hannahsonsalla/Desktop/Math/DataScience/Project/minnesotaXC.csv")
```

Clean data

Code for fixing MIAC conference championships 2013

miac_xc = read.csv("/Users/hannahsonsalla/Desktop/Math/DataScience/Project/miacCC171022.csv")
conference_2013 = read.csv("/Users/hannahsonsalla/Desktop/Math/DataScience/Project/conference2013.csv")

minnesota_xc <-
  miac_xc %>% 
  filter(!meet_name == "MIAC Cross Country Championships") %>% 
  rbind(conference_2013)

write.csv(minnesota_xc,"/Users/hannahsonsalla/Desktop/Math/DataScience/Project/miacCC171027.csv")

Read in CSV
```{r}
# corrected for augsburg meets, miac 2013
minnesota_xc = read.csv("/Users/hannahsonsalla/Desktop/Math/DataScience/Project/miacCC171027.csv")
```


Clean all of minnesota XC college data
```{r}
minnesota_xc_clean <- 
  minnesota_xc %>%

  # get rid of \n in columns
  mutate(meet_name = gsub("(\n)*","", meet_name)) %>%
  mutate(place = gsub("(\n)*","",place)) %>%
  mutate(name = gsub("(\n)*","", name)) %>%
  mutate(class_year = gsub("(\n)*","", class_year)) %>%
  mutate(team = gsub("(\n)*","", team)) %>%
  mutate(avg_mile = gsub("(\n)*","", avg_mile)) %>%
  mutate(time = gsub("(\n)*","", time)) %>%
  mutate(score = gsub("(\n)*","",score)) %>% 
    
  # get rid of leading and trailing whitespace 
  mutate(meet_name = trimws(meet_name, "both")) %>%
  mutate(meet_location = trimws(meet_location, "both")) %>% 

  # need to remove row with Div I West Qualifying header info
  filter(!grepl("2017", name)) %>% 
  # get rid of irrelevant header rows
  filter(place != "Place")  %>%
  filter(place != "Rank")  %>%
  filter(!grepl("\\d+", team)) %>% 
  filter(!grepl("\xfc\xbe\x8c\x86\x84\xbc",team))  %>%
  # filter out DNF runners 
  filter(time != "DNF") %>%
  # filter out runners with no name
  filter(name != ""  & name != " ") %>%
  # filter out unattached runners 
  filter(!grepl("Unattatched|Unattached|unattached", team)) %>% 
  # filter out alumni 
  filter(!grepl("Alumni|alumni|Alum", team)) %>%
  # filter out track clubs and Run n Fun
  filter(!grepl("club|Club", team)) %>%
  filter(!grepl("Run|Running", team)) %>%
  # get rid of Fall Finale 2013 meet - men and women results intermixed
  filter(!meet_name == "Fall Finale") %>% 
  
  # replace divs in score with NA
  mutate(score = ifelse(grepl("\xfc\xbe\x8c\x86\x84\xbc", score), NA, score)) %>%
  
  # shorten start time text
  mutate(start_time = gsub("(\n)*","", start_time)) %>%
  mutate(start_time = substr(start_time,start=13,stop=nchar(start_time))) %>%
  #filter rows with time more than 1 hour
  #filter(!grepl("?=:.*:", time)) %>% 
  
  # convert race date to date format
  mutate(meet_date = as.Date(meet_date, "%m/%d/%y")) %>% 
  # add meet year column
  mutate(meet_year = as.character(year(meet_date))) %>%
  
  # add column for meet distance
  mutate(meet_kilometers = ifelse(grepl("3k", race_group), 3, "NA")) %>%
  mutate(meet_kilometers = ifelse(grepl("3.2k", race_group), 3.2, meet_kilometers)) %>%
  mutate(meet_kilometers = ifelse(grepl("4k", race_group), 4, meet_kilometers)) %>%
  mutate(meet_kilometers = ifelse(grepl("5k", race_group), 5, meet_kilometers)) %>%
  mutate(meet_kilometers = ifelse(grepl("6k", race_group), 6, meet_kilometers)) %>%
  mutate(meet_kilometers = ifelse(grepl("6.437k|4 Mile", race_group), 6.437, meet_kilometers)) %>%
  mutate(meet_kilometers = ifelse(grepl("8k", race_group), 8, meet_kilometers)) %>%
  # alter race group to be of values men or women
  mutate(race_group = ifelse(grepl("Women", race_group), "Women", "Men")) %>%
    
  # only get location to first ,
  mutate(meet_location = gsub("(.*?)(,.*)", "\\1", meet_location )) %>%
  mutate(meet_location = gsub("(.*?)(-.*)", "\\1", meet_location )) %>%
  
  # condense teams
  mutate(team = ifelse(grepl("Augsburg", team), "Augsburg", team)) %>%
  mutate(team = ifelse(grepl("Bethel", team), "Bethel", team)) %>%
  mutate(team = ifelse(grepl("Carleton", team), "Carleton", team)) %>%
  mutate(team = ifelse(grepl("Concordia.*Moorhead|Moorhead.*Concordia", team), "Concordia Moorhead", team)) %>%
  mutate(team = ifelse(grepl("Gustavus Adolphus",team), "Gustavus Adolphus", team)) %>%
  mutate(team = ifelse(grepl("Hamline", team), "Hamline", team)) %>%
  mutate(team = ifelse(grepl("Macalester", team), "Macalester", team)) %>%
  mutate(team = ifelse(grepl("St. Benedict", team), "St. Benedict", team)) %>%
  mutate(team = ifelse(grepl("St. Catherine", team), "St. Catherine", team)) %>%
  mutate(team = ifelse(grepl("St. John's", team), "St. John's", team)) %>%
  mutate(team = ifelse(grepl("Saint Mary's", team), "Saint Mary's", team)) %>%
  mutate(team = ifelse(grepl("St. Olaf", team), "St. Olaf", team)) %>%
  mutate(team = ifelse(grepl("St. Thomas", team), "St. Thomas", team)) %>%
  
  # condense locations
  mutate(meet_location = ifelse(meet_location == "St. Kates", "St. Catherine", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("St. Catherine", meet_location), "St. Catherine", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("Carleton", meet_location), "Carleton", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("St. Olaf", meet_location), "St. Olaf", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("Como Golf Course", meet_location), "Como Park Golf Course", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("Mary's", meet_location), "Saint Mary's", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("Les Bolstad", meet_location), "Les Bolstad Golf Course", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("Gustavus Adolphus", meet_location), "Gustavus Adolphus", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("Gusvaus Adlophus", meet_location), "Gustavus Adolphus", meet_location)) %>%
  mutate(meet_location = ifelse(grepl("Crown", meet_location), "Crown", meet_location)) %>% 
    
  # separate time into minutes and seconds
  separate(time, c("race_minutes", "race_seconds"), sep = ":", remove = FALSE) %>% 
  mutate(new_time = as.numeric(race_minutes) + (as.numeric(race_seconds)/60)) 
```

Note: row 1821 and row 1822 have times over 1 hour (1:07:00.5 and 1:07:43.9)

MIAC XC data

MIAC colleges: Augsburg University, Bethel University, Carleton College, Concordia Moorhead College, Gustavus Adolphus College, Hamline University, Macalester College, College of Saint Benedict, St. Catherine University, Saint John's University, Saint Mary's University, St. Olaf College, University of St. Thomas 
```{r}
miacSchools = c("Augsburg", "Bethel", "Carleton", "Concordia Moorhead", "Gustavus Adolphus", "Hamline", "Macalester", "St. Benedict", "St. Catherine", "St. John's", "Saint Mary's", "St. Olaf", "St. Thomas")

miac_xc <-
  minnesota_xc_clean %>%
  filter(team %in% miacSchools) %>% 
  # MIAC Championships 2014 has two different links
  filter(!grepl("https://www.tfrrs.org/results/xc/7815.html", url))
```

Let's look team size by school in the MIAC for 2017.

Men:
```{r}
miac_team_size_men <-
  miac_xc %>%
  filter(race_group == "Men") %>% 
  distinct(name, meet_year, team, .keep_all = TRUE)
```

Stacked Plot - to compare which school attracts most runners
```{r}
team_size_men_plot1 <- ggplot(miac_team_size_men, aes(team)) + geom_bar(aes(fill = meet_year)) + labs(title = "MIAC Men Cross Country Team Size", subtitle = "Years 2012 - 2017", x = "College/University",  y = "number of runners", fill = "year") +  theme(axis.text.x=element_text(angle=45,hjust=1)) + scale_fill_manual(values = c("red","salmon", "pink","slategray1", "steelblue1","steelblue4"))

team_size_men_plot1
```

Side-by-Side Plot - to look at which schools have largest teams per year
```{r}
team_size_plot_men2 <- ggplot(miac_team_size_men, aes(team)) + geom_bar(aes(fill = meet_year), position = "dodge") + labs(title = "MIAC Men Cross Country Team Size", subtitle = "Years 2012 - 2017", x = "College/University",  y = "number of runners", fill = "year") +  theme(axis.text.x=element_text(angle=45,hjust=1)) + scale_fill_manual(values = c("red","salmon", "pink","slategray1", "steelblue1","steelblue4"))

team_size_plot_men2
```

Women:
```{r}
miac_team_size_women <-
  miac_xc %>%
  filter(race_group == "Women") %>% 
  distinct(name, meet_year, team, .keep_all = TRUE)
```

Stacked Plot - to compare which school attracts most runners
```{r}
team_size_women_plot1 <- ggplot(miac_team_size_women, aes(team)) + geom_bar(aes(fill = meet_year)) + labs(title = "MIAC Women Cross Country Team Size", subtitle = "Years 2012 - 2017", x = "College/University",  y = "number of runners", fill = "year") +  theme(axis.text.x=element_text(angle=45,hjust=1)) + scale_fill_manual(values = c("red","salmon", "pink","slategray1", "steelblue1","steelblue4"))

team_size_women_plot1
```

Side-by-Side Plot - to look at which schools have largest teams per year
```{r}
team_size_women_plot2 <- ggplot(miac_team_size_women, aes(team)) + geom_bar(aes(fill = meet_year), position = "dodge") + labs(title = "MIAC Women Cross Country Team Size", subtitle = "Years 2012 - 2017", x = "College/University",  y = "number of runners", fill = "year") +  theme(axis.text.x=element_text(angle=45,hjust=1)) + scale_fill_manual(values = c("red","salmon", "pink","slategray1", "steelblue1","steelblue4"))

team_size_women_plot2
```

MIAC Championships - Winner, All Conference, Honorable Mention by Year

Men
```{r}
places = c("1", "15", "25")
miac_men_xc_championships <-
  miac_xc %>% 
  filter(race_group == "Men") %>% 
  filter(grepl("MIAC", meet_name)) %>%
  filter(place %in% places) %>%
  mutate(place_category = ifelse(as.numeric(place) == 15, "All Conference", "Honorable Mention")) %>% 
  mutate(place_category = ifelse(as.numeric(place) == 1, "Champion", place_category))  
miac_men_xc_championships2 <-
  miac_xc %>%
  filter(race_group == "Men") %>% 
  filter(grepl("MIAC", meet_name)) %>%
  filter(as.numeric(place) < 26) %>%
  mutate(place_category = ifelse(as.numeric(place) <= 15, "All Conference", "Honorable Mention")) %>%
  mutate(place_category = ifelse(as.numeric(place) == 1, "Champion", place_category)) 
```

Plot
```{r}
miac_men_conference_place_plot <- ggplot(miac_men_xc_championships, aes(x = meet_year, y = new_time, group = place_category)) + geom_point(data = miac_men_xc_championships2, alpha = 0.5, aes(color = place_category)) + geom_point(aes(color = place_category), size = 3) + geom_line(aes(color = place_category)) + scale_color_manual(values = c("red", "green", "blue"), breaks=c("Champion","All Conference","Honorable Mention")) + labs(title = "MIAC Championships Men (8k)", subtitle = "Champion, All-Conference, Honorable Mention", x = "year",  y = "time (minutes)", fill = "place")

miac_men_conference_place_plot
```


Women
```{r}
places = c("1", "15", "25")
miac_women_xc_championships <-
  miac_xc %>% 
  filter(race_group == "Women") %>% 
  filter(grepl("MIAC", meet_name)) %>%
  filter(place %in% places) %>%
  mutate(place_category = ifelse(as.numeric(place) == 15, "All Conference", "Honorable Mention")) %>% 
  mutate(place_category = ifelse(as.numeric(place) == 1, "Champion", place_category))  
miac_women_xc_championships2 <-
  miac_xc %>%
  filter(race_group == "Women") %>% 
  filter(grepl("MIAC", meet_name)) %>%
  filter(as.numeric(place) < 26) %>%
  mutate(place_category = ifelse(as.numeric(place) <= 15, "All Conference", "Honorable Mention")) %>%
  mutate(place_category = ifelse(as.numeric(place) == 1, "Champion", place_category)) 
```

Plot
```{r}
miac_women_conference_place_plot <- ggplot(miac_women_xc_championships, aes(x = meet_year, y = new_time, group = place_category)) + geom_point(data = miac_women_xc_championships2, alpha = 0.5, aes(color = place_category)) + geom_point(aes(color = place_category), size = 3) + geom_line(aes(color = place_category)) + scale_color_manual(values = c("red", "green", "blue"), breaks=c("Champion","All Conference","Honorable Mention")) + labs(title = "MIAC Championships Women (6k)", subtitle = "Champion, All Conference, Honorable Mention", x = "year",  y = "time (minutes)", fill = "place")

miac_women_conference_place_plot
```

MIAC Courses 2016

Men
```{r}
miac_men_courses_2016 <-
  miac_xc %>% 
  filter(race_group == "Men") %>%
  filter(meet_year == 2016) %>%
  separate(avg_mile, c("mile_minutes", "mile_seconds"), sep = ":", remove = FALSE) %>% 
  mutate(new_mile = as.numeric(mile_minutes) + (as.numeric(mile_seconds)/60)) %>% 
  mutate(meet_name = gsub("Cross Country Championships", "", meet_name))

miac_men_courses_2016$meet_name <- factor(miac_men_courses_2016$meet_name, levels = miac_men_courses_2016$meet_name[order(miac_men_courses_2016$meet_date)])
```

Plot
```{r}
men_courses_2016_plot <- ggplot(miac_men_courses_2016, aes(x = meet_name, y = new_mile)) + geom_point(alpha = 1/10) + geom_boxplot() + labs(title = "MIAC Men Cross Country Courses 2016", x = "courses",  y = "average mile (minutes)") +  theme(axis.text.x=element_text(angle=45,hjust=1))

men_courses_2016_plot
```

Women
```{r}
miac_women_courses_2016 <-
  miac_xc %>% 
  filter(race_group == "Women") %>%
  filter(meet_year == 2016) %>%
  separate(avg_mile, c("mile_minutes", "mile_seconds"), sep = ":", remove = FALSE) %>% 
  mutate(new_mile = as.numeric(mile_minutes) + (as.numeric(mile_seconds)/60)) %>% 
  mutate(meet_name = gsub("Cross Country Championships", "", meet_name))

miac_women_courses_2016$meet_name <- factor(miac_women_courses_2016$meet_name, levels = miac_women_courses_2016$meet_name[order(miac_women_courses_2016$meet_date)])
```

Plot

Had to use race time instead of average mile.  St. Olaf Invitational calculated mile time in terms of 8k instead of 6k for women so it's incorrect.
```{r}
women_courses_2016_plot <- ggplot(miac_women_courses_2016, aes(x = meet_name, y = new_time)) + geom_point(alpha = 1/10) + geom_boxplot() + labs(title = "MIAC Women Cross Country Courses 2016", x = "courses",  y = "average time (minutes)") +  theme(axis.text.x=element_text(angle=45,hjust=1))

women_courses_2016_plot
```

Progression of 4 year runners

Men
```{r}
four_years_miac_men <-
  miac_xc %>% 
  filter(race_group == "Men") %>%
  filter(meet_kilometers == 8) %>%
  group_by(name, meet_year) %>%
  summarise(year_pr = min(new_time)) %>%
  # want years 2013 - 2016
  filter(!meet_year == 2012 & !meet_year == 2017) %>%
  group_by(name) %>%
  filter(n() == 4)
```

Plot of all MIAC men four year runners 2013 - 2016
```{r}
miac_four_years_men_plot <- ggplot(four_years_miac_men, aes(x = meet_year, y = year_pr)) + geom_point(alpha = 1/2) + labs(title = "MIAC Four Year Runners", subtitle = "Years 2013 - 2016", x = "year",  y = "time (minutes)") + geom_boxplot()

miac_four_years_men_plot
```

Individual 3 runners
```{r}
runners <- "Krogman, Brandon|Edstrom, Walter|Ryan, Mike"
men_four_runners_three <-
  four_years_miac_men %>%
  filter(grepl(runners, name))
```

Plot
```{r}
miac_four_years_men_plot2 <- ggplot(men_four_runners_three, aes(x = meet_year, y = year_pr, group = name)) + geom_point(aes(color = name), size = 2) + geom_line(aes(color = name)) + labs(title = "MIAC Men Four Year Runners", subtitle = "Years 2013 - 2016", x = "year",  y = "time (minutes)") 

miac_four_years_men_plot2
```