---
title: "R Notebook"
output: html_notebook
---

Research Questions:
1. How much recruiting/interactions with recruits do coaches need to do to get a full roster each year?
2. What does the current athletic student body look like, could this help you target recruits?
3. Once a recruited student has been accepted what is the rate for which students attend vs. not and what was different about their experiences that may have influenced their decision.

1. Data from athletics
2. Data from admissions?
3. Historical roster lists
4. Data interest form online could be helpful

```{r}
library(rvest)
library(dplyr)
library(magrittr)
library(tidyverse)
library(jsonlite)
library(stringr)
library(xml2)
library(XML)
library(rvest)
library(tidyverse)
library(httr)
library(httpuv)
library(RCurl)
```

This chunk of code didn't work well, use second chunk
```{r}
athleteData <- NULL
athleteData2 <- NULL
df <- NULL
df2 <- NULL
flag <- 0
flag2 <- 0

#for(i in 1:411) {
 i=1
  url = paste("http://athletics.macalester.edu/roster.aspx?roster=",i,sep="")
  teamPage <- read_html(url)
  
  teamYearRoster <- teamPage %>% 
    html_node(".page_title") %>%
    html_nodes("h2") %>% 
    html_text() %>% 
    toString()
  
  rosterList <- teamPage %>% 
    html_nodes(".roster_dgrd_item") #(".default_dgrd_alt"))
  
  if(length(rosterList) > 0) {
    for(j in 1:length(rosterList)) {
      name = ""
      year = ""
      hometownHS = ""
      
      name <- rosterList[j] %>% 
        html_node(".roster_dgrd_full_name") %>% 
        html_node("a") %>% 
        html_text() %>% 
        toString()
      
      year <- rosterList[j] %>% 
        html_node(".roster_dgrd_academic_year") %>% 
        html_text() %>% 
        toString()
            
      location <- rosterList[j] %>% 
        html_node(".roster_dgrd_hometownhighschool") %>% 
        html_text() %>% 
        toString()
      
        df <- data.frame(teamYearRoster,name,year,location,i)
  
      if(flag == 0) {
        athleteData <- df
      } else {
        athleteData <- rbind(athleteData,df)
      }
      
      flag <- 1
      
    }
  }
  
  rosterList2 <- teamPage %>% 
    html_nodes(".default_dgrd_alt")
  
  if(length(rosterList2) > 0) {
    for(j in 1:length(rosterList2)) {
      name = ""
      year = ""
      hometownHS = ""
      
      name <- rosterList2[j] %>% 
        html_node(".roster_dgrd_full_name") %>% 
        html_node("a") %>% 
        html_text() %>% 
        toString()
      
      year <- rosterList2[j] %>% 
        html_node(".roster_dgrd_academic_year") %>% 
        html_text() %>% 
        toString()
            
      location <- rosterList2[j] %>% 
        html_node(".roster_dgrd_hometownhighschool") %>% 
        html_text() %>% 
        toString()
      
        df2 <- data.frame(teamYearRoster,name,year,location,i)
  
      if(flag2 == 0) {
        athleteData2 <- df2
      } else {
        athleteData2 <- rbind(athleteData2,df2)
      }
      
      flag2 <- 1
      
    }
  }
  
  print(url)
  
#}
  
finalAthleteRoster <- rbind(athleteData,athleteData2)
  
#teamPage <- read_html("http://athletics.macalester.edu/roster.aspx?roster=1")
```
#flatten <- will flatten nested data frames into it's own columns basically
```{r}
athleteData <- NULL
df <- NULL
flag <- 0

for(i in 1:411) {
  #i=1
  url = paste("http://athletics.macalester.edu/roster.aspx?roster=",i,sep="")
  teamPage <- read_html(url)
  
  teamYearRoster <- teamPage %>% 
    html_node(".page_title") %>%
    html_nodes("h2") %>% 
    html_text() %>% 
    toString()
  
  nameList <- teamPage %>% 
    html_nodes(".roster_dgrd_full_name")
    
  yearList <- teamPage %>% 
    html_nodes(".roster_dgrd_academic_year")
  
  hometownList <- teamPage %>% 
    html_nodes(".roster_dgrd_hometownhighschool")
  
  if(length(nameList) > 0) {
    for(j in 1:length(nameList)) {
      name = ""
      year = ""
      hometownHS = ""
      
      name <- nameList[j] %>% 
        html_node("a") %>% 
        html_text() %>% 
        toString()
      
      year <- yearList[j] %>% 
        html_text() %>% 
        toString()
            
      location <- hometownList[j] %>% 
        html_text() %>% 
        toString()
      
        df <- data.frame(teamYearRoster,name,year,location,i)
  
      if(flag == 0) {
        athleteData <- df
      } else {
        athleteData <- rbind(athleteData,df)
      }
      
      flag <- 1
      
    }
  }
  print(url)
}

#write.csv(athleteData,"C:/Users/CPC24/Documents/HeatherRTrainingFun/athleteData171004.csv")
```
```{r}
athleteData <- athleteData %>% 
  mutate()
```


```{r}
athleteData <- read_csv("C:/Users/CPC24/Documents/HeatherRTrainingFun/athleteData171004.csv")

athleteDataClean <- athleteData %>% 
  separate(location,c("citystate","highschool"),sep=" / ") %>% 
  separate(citystate,c("city","state"),sep=", ")

cleanState <- function(state){
  ifelse((state=="Ak"|state=="AK"),"AK",
         ifelse((state=="AL"),"AL",
                ifelse((state=="AL"|state=="Ala."|state=="Alab."),"AL")))
}


states <- athleteDataClean %>% distinct(state) %>% arrange(state)
```

```{r}
#Scrape the data from this page: https://www.tfrrs.org/results_search.html?page=0&title=1&go=1&state=mn&sport=xc to get links to the individual result links then scrap those

flag <- 0
df <- NULL
resultLinks <- NULL
resultList <- read_html("https://www.tfrrs.org/results_search.html?page=0&title=1&go=1&state=mn&sport=xc")

links <- resultList %>% 
  html_nodes("#results_search a")


for(i in 1:length(links)) {
  
  link <- links[i] %>% 
    html_attrs() %>% 
    toString()
 
  df <- tibble(link)

  if(flag == 0) {
    resultLinks <- df
  } else {
    resultLinks <- rbind(resultLinks,df)
  }
  
  flag <- 1
   
}

resultLinksClean <- resultLinks %>% 
  mutate(link = gsub("(//www)","https://www",link),
         link = as.character(link))

write.csv(resultLinks,"/Users/hannahsonsalla/Desktop/resultLinks.csv")

```

Final
```{r}
resultLinks <- read.csv("/Users/hannahsonsalla/Desktop/resultLinks.csv")

resultLinks <- resultLinks %>%
  mutate(link = gsub("(//www)","https://www",link),
         link = as.character(link))

miac_women_xc<-data.frame()

#miac_women_xcBU <- miac_women_xc

# meet urls

# ***table:nth-child(17)*** 
#Carleton Running of the Cows 2017 - https://www.tfrrs.org/results/xc/12950.html 

# ***table:nth-child(13) ***
# St. Kates Invite 2017 - https://www.tfrrs.org/results/xc/13115.html
# Having trouble with alum team
# Roy Griak 2017 - https://www.tfrrs.org/results/xc/12864.html
# Having trouble with Lakeland University

# *** table:nth-child(6) *** 
# Alum individual
# Bob Waxlax Invitational 2017 - https://www.tfrrs.org/results/xc/13109.html

# *** table:nth-child(8) *** 
# St. Olaf Invitational 2017 - https://www.tfrrs.org/results/xc/12550.html



# loop for table:nth-child(17) 
for (url in resultLinks$link) {
  
    #url=resultLinks$link[1]
    #url = "https://www.tfrrs.org/results/xc/11018.html"
    #read the page 
    page<- read_html(url)
    
        #Retrieve meet name
    meet_name <-
      page %>%
      html_nodes(".title a") %>%
      html_text()
    
    #Retrieve meet date
    meet_date <-
      page %>%
      html_nodes(".datelocation li:nth-child(1)") %>%
      html_text()
    meet_date_short<-substr(meet_date,start=6,stop=14)
    
    #Retrieve meet location
    meet_location <-
      page %>%
      html_nodes(".datelocation li:nth-child(2)") %>%
      html_text()
    meet_location_short<-substr(meet_location,start=12,stop=(nchar(meet_location) - 4))
    meet_location_short = gsub("(\n)|(\nCity Not Supplied,)",", ",meet_location_short)
    
    start_time <-
      page %>%
      html_node("div:nth-child(12) span") %>%
      html_text()
    start_time = gsub("(\n)*(\t)*","",start_time)
    
    race <- page %>% 
      html_nodes("a+ div div")
    
    #Last table on page contains womens results
    tables <- page %>% 
      html_nodes("table")
    
    lastTableLocation <- length(tables)
    
    tableList = c(lastTableLocation)
    tableRef = lastTableLocation - 2
    while(tableRef >= 4) {
      tableList=c(tableRef,tableList)
      tableRef = tableRef - 2
    }
    
     
    for(t in (1:length(tableList))) { 
      
      race_group=""
      if (length(race[t] %>% html_text()>0)) {
        race_group = race[t] %>% html_text()
        race_group = gsub("(\n)*(\t)*","",race_group)
      } 
    
      lastTableDivs <- tables[tableList[t]] %>% 
        html_nodes("tr") %>% 
        html_nodes("td") %>% 
        html_nodes("div")
      
      #Need to change this to only loop through first 7 td divs in the table
      tableRows <- tables[tableList[t]] %>% 
        html_nodes("tr")
      
      for(row in (1:length(tableRows))) {
        
        tableDivs <- tableRows[row] %>% 
          html_nodes("td")  %>% 
          html_nodes("div")
        
        if(length(tableDivs)>0) {
        
         # Retrieve place
          women_place <- tableDivs[1] %>%
            html_text()
          
          # Retrieve and inspect full names
          women_name <- tableDivs[2] %>%
            html_text()
          women_name_short<-substr(women_name,start=10,stop=nchar(women_name))
      
          # Retrieve and inspect class year
          women_class_year <- tableDivs[3] %>%
            html_text()
      
          # Retrieve and inspect team
          women_team <- tableDivs[4] %>%
            html_text()
          women_team_short<-substr(women_team,start=10,stop=nchar(women_team))
      
          # Retrieve and inspect avg. mile
          women_avg_mile <- tableDivs[5] %>%
            html_text()
          
          # Retrieve time
          women_time <- tableDivs[6] %>%
            html_text()
          
          # Retrieve score
          women_score <- tableDivs[7] %>%
            html_text()
          
          
          meet_info <- 
            tibble(meet_name = meet_name, meet_date = meet_date_short, meet_location = meet_location_short, start_time=start_time, race_group=race_group, place = women_place, name = women_name_short, class_year = women_class_year, team = women_team_short, avg_mile = women_avg_mile, time = women_time, score = women_score, url=url)
        
          #store the temp values into the master dataframe
          miac_women_xc<-rbind(miac_women_xc, meet_info)
        }
      }
    }
    print(url)
} 

head(miac_women_xc)

write.csv(miac_women_xc,"/Users/hannahsonsalla/Desktop/miacCCResults.csv")

test <- miac_women_xc %>% 
  mutate(race_group = gsub("(\n)*(\t)*","",race_group),
         start_time = gsub("(\n)*(\t)*","",start_time),
         meet_location = gsub("(\n)|(\nCity Not Supplied,)",", ",meet_location)) %>% 
  filter(place!="Place")

```

Add get "div" get "a" and add the link to show which is mens and which is womens since the order changes


```{r}
##ORIGINAL
# loop for table:nth-child(17) 
for (url in resultLinks$link) {
  
    #url=resultLinks$link[1]
    #url = "https://www.tfrrs.org/results/xc/12864.html"
    #read the page 
    page<- read_html(url)
    
        #Retrieve meet name
    meet_name <-
      page %>%
      html_nodes(".title a") %>%
      html_text()
    
    #Retrieve meet date
    meet_date <-
      page %>%
      html_nodes(".datelocation li:nth-child(1)") %>%
      html_text()
    meet_date_short<-substr(meet_date,start=6,stop=14)
    
    #Retrieve meet location
    meet_location <-
      page %>%
      html_nodes(".datelocation li:nth-child(2)") %>%
      html_text()
    meet_location_short<-substr(meet_location,start=12,stop=(nchar(meet_location) - 4))
    
    start_time <-
      page %>%
      html_node("div:nth-child(12) span") %>%
      html_text()
    
    race <- page %>% 
      html_nodes("a+ div div")
    
    #Last table on page contains womens results
    tables <- page %>% 
      html_nodes("table")
    
    lastTableLocation <- length(tables)
    
    tableList = c(lastTableLocation)
    tableRef = lastTableLocation - 2
    while(tableRef >= 4) {
      tableList=c(tableRef,tableList)
      tableRef = tableRef - 2
    }
    
     
    for(t in (1:length(tableList))) { 
      
      race_group=""
      if (length(race[t] %>% html_text()>0)) {
        race_group = race[t] %>% html_text()
      } 
    
      lastTableDivs <- tables[tableList[t]] %>% 
        html_nodes("tr") %>% 
        html_nodes("td") %>% 
        html_nodes("div")
      
      #Need to change this to only loop through first 7 td's
      tableTDs <- tables[tableList[t]] %>% 
        html_nodes("tr") %>% 
        html_nodes("td") 
      
      tableDivs <- tableTDs %>% 
        html_nodes("div")
      
      for(divCol in (1:7))
      ##
      
      #will need to remove header rows but this is easiest.
      for(y in 1:(length(lastTableDivs)/7)-1) {
        
         # Retrieve place
          women_place <- lastTableDivs[1+7*y] %>%
            html_text()
          
          # Retrieve and inspect full names
          women_name <- lastTableDivs[2+7*y] %>%
            html_text()
          women_name_short<-substr(women_name,start=10,stop=nchar(women_name))
      
          # Retrieve and inspect class year
          women_class_year <- lastTableDivs[3+7*y] %>%
            html_text()
      
          # Retrieve and inspect team
          women_team <- lastTableDivs[4+7*y] %>%
            html_text()
          women_team_short<-substr(women_team,start=10,stop=nchar(women_team))
      
          # Retrieve and inspect avg. mile
          women_avg_mile <- lastTableDivs[5+7*y] %>%
            html_text()
          
          # Retrieve time
          women_time <- lastTableDivs[6+7*y] %>%
            html_text()
          
          # Retrieve score
          women_score <- lastTableDivs[7+7*y] %>%
            html_text()
          
          
          meet_info <- 
            tibble(meet_name = meet_name, meet_date = meet_date_short, meet_location = meet_location_short, start_time=start_time, race_group=race_group, place = women_place, name = women_name_short, class_year = women_class_year, team = women_team_short, avg_mile = women_avg_mile, time = women_time, score = women_score, url=url)
        
          #store the temp values into the master dataframe
          miac_women_xc<-rbind(miac_women_xc, meet_info)
          
      }
    }
    print(url)
    print(y)
} 

```

