---
title: "MACAthletics"
output: html_document
---

Recruiting Research Questions:
1. How much recruiting/interactions with recruits do coaches need to do to get a full roster each year?
2. What does the current athletic student body look like, could this help you target recruits?
3. Once a recruited student has been accepted what is the rate for which students attend vs. not and what was different about their experiences that may have influenced their decision.

Data Aquisition
1. Data from athletics
2. Data from admissions?
3. Historical roster lists
4. Data interest form online could be helpful

```{r}
library(rvest)
library(dplyr)
library(magrittr)
library(tidyverse)
library(jsonlite)
library(stringr)
library(xml2)
library(XML)
library(rvest)
library(tidyverse)
library(httr)
library(httpuv)
library(RCurl)
```

Data scraping
Scrape data from MAC rosters online.  Each row represents an athlete.  
```{r}
macAthleteData <- data.frame()

for(i in 1:411) {
  #i=1
  url = paste("http://athletics.macalester.edu/roster.aspx?roster=",i,sep="")
  teamPage <- read_html(url)
  
  teamYearRoster <- teamPage %>% 
    html_node(".page_title") %>%
    html_nodes("h2") %>% 
    html_text() %>% 
    toString()
  
  nameList <- teamPage %>% 
    html_nodes(".roster_dgrd_full_name")
    
  yearList <- teamPage %>% 
    html_nodes(".roster_dgrd_academic_year")
  
  hometownList <- teamPage %>% 
    html_nodes(".roster_dgrd_hometownhighschool")
  
  if(length(nameList) > 0) {
    for(j in 1:length(nameList)) {
      name = ""
      year = ""
      hometownHS = ""
      
      name <- nameList[j] %>% 
        html_node("a") %>% 
        html_text() %>% 
        toString()
      
      year <- yearList[j] %>% 
        html_text() %>% 
        toString()
            
      location <- hometownList[j] %>% 
        html_text() %>% 
        toString()
      
      athleteInfo <- tibble(sportYear = teamYearRoster,name = name,year = year, hometown = location,i)
      macAthleteData<-rbind(macAthleteData, athleteInfo)
    }
  }
  print(url)
}
write.csv(macAthleteData,"/Users/hannahsonsalla/Desktop/Math/DataScience/Project/macAthleteData.csv")
```

Clean MAC athlete data set

Separate hometown into city, state and highschool columns.
```{r}
macAthleteData <- read.csv("/Users/hannahsonsalla/Desktop/Math/DataScience/Project/macAthleteData.csv")

macAthleteDataClean <- macAthleteData %>% 
  separate(hometown,c("citystate","highschool"),sep=" / ") %>% 
  separate(citystate,c("city","state"),sep=", ")
```

Let's look at the cross country rosters and see how team size has changed over time.

Filter for women cross country athletes
```{r}
macXC <-
  macAthleteDataClean %>%
  # filter for cross country
  filter(grepl("Cross Country", sportYear))  %>%
  #separate sportYear into academicYear and sport
  separate(sportYear, into = c('academicYear', 'sport'), sep = 4) %>%
  # add gender column
  mutate(gender = ifelse(grepl("Women", sport), "Women", "Men")) %>%
  # eliminate Macalester, Roster, Women's, Men's from sport column
  mutate(sport = gsub("Macalester", "", sport)) %>%
  mutate(sport = gsub("Roster", "", sport)) %>%
  mutate(sport = gsub("Men's", "", sport), sport = gsub("Women's", "", sport))
```

Macaleser Cross Country Plot
```{r}
xcPlot <- ggplot(macXC, aes(x = academicYear)) + geom_bar(fill = "steelblue4") + theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(y = "number of runners", x = "year", title = "Macalester College Cross Country", subtitle = "Men's and Women's Programs")

xcPlot
```


Macaleser Cross Country Plot by Gender
```{r}
xcPlot2 <- ggplot(macXC, aes(x = academicYear)) + geom_bar(aes(fill = gender)) + theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(y = "number of runners", x = "year", title = "Macalester College Cross Country", subtitle = "Men's and Women's Programs") + scale_fill_manual(values=c("chocolate", "steelblue4"))

xcPlot2
```

Macaleser Women's Cross Country
```{r}
macWomenXC <- 
  macXC %>%
  filter(gender == "Women")
```

Plot of Team Size by Year
```{r}
xcPlot3 <- ggplot(macWomenXC, aes(x = academicYear)) + geom_bar(fill = "chocolate") +  theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(y = "number of runners", x = "year", title = "Macalester College Women's Cross Country", subtitle = "Team Size by Year")

xcPlot3
```

Plot by Class Year
```{r}
xcPlot4 <- ggplot(macWomenXC, aes(x = academicYear)) + geom_bar(aes(fill = year)) + theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(y = "number of runners", x = "year", title = "Macalester College Women's Cross Country", subtitle = "Team Composition by Class Year", fill = "class year") + scale_fill_manual(values = c("red","salmon","steelblue1","steelblue4"))

xcPlot4
```


Macaleser Men's Cross Country 
```{r}
macMenXC <- 
  macXC %>%
  filter(gender == "Men")
```

Plot of Team Size by Year
```{r}
xcPlot5 <- ggplot(macMenXC, aes(x = academicYear)) + geom_bar(fill = "steelblue4") +  theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(y = "number of runners", x = "year", title = "Macalester College Men's Cross Country", subtitle = "Team Size by Year")

xcPlot5
```

Plot by Class Year
```{r}
xcPlot6 <- ggplot(macMenXC, aes(x = academicYear)) + geom_bar(aes(fill = year)) + theme(axis.text.x=element_text(angle=45,hjust=1)) + labs(y = "number of runners", x = "year", title = "Macalester College Men's Cross Country", subtitle = "Team Composition by Class Year", fill = "class year") + scale_fill_manual(values = c("red","salmon","steelblue1","steelblue4"))

xcPlot6
```


